The CrypTool Script: \\ Cryptography, Mathematics and More \\ \vspace{30pt} {\large Background reading\\%*[-8pt] for the free e-learning program CrypTool\\ (10th edition -- distributed with CrypTool version 1.4.30)} \vspace{60pt}

### 2009-08-17 script/en/chapters/paper_and_pencil.tex
CT script, Appendixes in chap 2, 3 and 4 named in the same way (de+en)
fixed math mode in headings resulting in broken PDF bookmarks with \texorpdfstring; fixed float warnings by replacing placement [h] with [ht]; fixed warning about PDF version 1.5 by changing it to 1.4 in coverv1_*.pdf
CT script: Sage parts finished in de + en; several corrections
CT script: Minor corrections, de + en; final now for Beta04

\chapter{Paper and Pencil Encryption Methods}
\label{Kapitel_PaperandPencil}
(Christine St\"otzel, April 2004; Updates: B.+C. Esslinger, June 2005; Updates Minh Van Nguyen and B. Esslinger, July 2009)
\index{Paper- and pencil methods}

@@

The first paper- and pencil methods already arose about 3000 years ago, but new
procedures were developed during the past century, too. All paper and pencil 
methods are a matter of symmetric methods\index{Encryption!symmetric}. 
Even the earliest encryption algorithms use the basic principles such as 
transposition, substitution, block construction and their combinations. 
Hence it is worthwhile to closely consider this ``ancient'' methods especially
under didactic aspects.

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}}
	  & n &   & x &   & m &   & l &   & o &   & t &   & a &   & s &   & o &   & i &   & i &   & n \\
	a &   & e &   & a &   & p &   & e &   & f &   & r &   & n &   & p &   & s &   & t &   & o &   \\
\end{tabular}
\caption{Rail Fence cipher}
\end{center} 
\end{table}

@@

   In the example for a turning grille in the following table you can write
   4 times 16 characters of the cleartext on a piece of paper:
\begin{table}[h]
\begin{center}
\begin{tabular}{|cccc|cccc|}
\hline 	
	O & - & - & - & - & O & - & - \\
	- & - & - & O & O & - & - & O \\
	- & - & - & O & - & - & O & - \\
	- & - & O & - & - & - & - & - \\
\hline 	
	- & - & - & - & O & - & - & - \\
	O & - & O & - & - & - & O & - \\
	- & O & - & - & - & - & - & O \\
	- & - & - & O & O & - & - & - \\
\hline
\end{tabular}  
\caption{8x8 turning grille} 
\end{center}   
\end{table}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline 	
	K & E & Y \\
\hline
	a & n & e \\
	x & a & m \\
	p & l & e \\
	o & f & t \\
	r & a & n \\
	s & p & o \\
	s & i & t \\
	i & o & n \\
\hline
\end{tabular}
\caption{Simple columnar transposition}
\end{center} 
\end{table}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline 	
	C & O & L & U & M & N\\
	1 & 5 & 2 & 6 & 3 & 4\\
\hline
	a &   &   &   &   &  \\
	n & e & x &   &   &  \\
	a & m & p & l & e &  \\
	o & f & t & r & a & n\\
	s & p &   &   &   &  \\
	o & s & i & t &   &  \\
	i & o & n &   &   &  \\
\hline
\end{tabular}
\caption{Columnar transposition (General Luigi Sacco)}
\end{center} 
\end{table}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|ccccc||cc|ccccc|}
\hline 	
	  & W & O & R & D & S &   &   & D & O & R & S & W\\
\hline
	W & a & n & e & x & a &   & D & s & p & o & i & s\\
	O & m & p & l & e & o &   & O & e & p & l & o & m\\
	R & f & t & r & a & n &   & R & a & t & r & n & f\\
	D & s & p & o & s & i &   & S & n & i & o & - & t\\
	S & t & i & o & n & - &   & W & x & n & e & a & a\\
\hline
\end{tabular}  
\caption[Nihilist transposition]{Nihilist transposition\footnotemark}
\end{center} 
\end{table}

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|ccc|ccc|ccc|}
\hline 	
	  & K & {\bf E} & Y & {\bf E} & K & Y & {\bf E} & K & Y\\
\hline
	A & c & a & d & a & c & d & {\bf s} & a & a\\
	D & e & n & u & n & e & u & s & r & p\\
	X & s & i & s & i & s & s & i & f & i\\
	K & a & f & o & f & {\bf a} & o & u & l & o\\
	C & r & m & o & m & r & o & n & n & s\\
	W & f & c & o & c & f & o & k & t & g\\
	N & l & u & m & u & l & m & w & n & e\\
	S & n & a & r & a & n & r & d & o & o\\
	Y & t & r & a & r & t & {\bf a} & a & t & d\\
	{\bf E} & n & {\bf s} & p & {\bf s} & n & p & n & n & u\\
	D & o & s & i & s & o & i & i & i & s\\
	T & t & i & o & i & t & o & f & a & o\\	
	U & n & u & s & u & n & s & m & y & o\\
	B & i & n & g & n & i & g & c & r & o\\
	R & a & k & e & k & a & e & u & c & m\\
	G & y & w & o & w & y & o & a & e & r\\
	H & r & d & - & d & r & - & r & s & -\\
\hline
\end{tabular}  
\caption[Cadenus]{Cadenus\footnotemark}
\label{Cadenus-table-reference}
\end{center} 
\end{table}%

@@

   \begin{table}[h]
   \begin{center}
   \begin{tabular}{|c|cccccccccc|}
   \hline 	
	  & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9\\
   \hline
	  & K & - & - & E & Y & W & O & R & D & A\\
	1 & B & C & F & G & H & I & J & L & M & N\\
	2 & P & Q & S & T & U & V & X & Z & . & /\\
   \hline
   \end{tabular}  
   \caption{Straddling checkerboard with password ``Keyword''}
   \label{Straddling-Checkerboard-table-reference}
   \end{center} 
   \end{table}

@@

   \begin{table}[h]
   \begin{center}
   \begin{tabular}{|c|ccccc|}
   \hline
        message              &  F   &   I   &   G   &   H   &   T     \\
   \hline
	ciphertext           & {\tt 00101} & {\tt 01000} & {\tt 00110} & {\tt 00111} & {\tt 10011} \\
	unsuspicious message & {\tt itisw} & {\tt arman} & {\tt thesu} & {\tt nissh} & {\tt ining} \\
   \hline 
        Baconian Cipher      & {\tt itIsW} & {\tt aRman} & {\tt thESu} & {\tt niSSH} & {\tt IniNG} \\
   \hline
   \end{tabular}  
   \caption{Baconian cipher}
   \label{Baconian-table-reference}
   \end{center} 
   \end{table}

@@

   \begin{table}[h]
   \begin{center}
   \begin{tabular}{|c|c|c|c|c|}
   \hline
	K & E & Y & W & O\\
   \hline
	R & D & A & B & C\\
   \hline
	F & G & H & I & L\\
   \hline
	M & N & P & Q & S\\
   \hline
	T & U & V & X & Z\\
   \hline
   \end{tabular}
   \caption{5x5 Playfair matrix}
   \label{Playfair-table-reference}
   \end{center}
   \end{table}

@@

   \begin{table}[h]
   \begin{center}
   \begin{tabular}{|ccccc|ccccc|}
   \hline
	d & w & x & y & m & E & P & T & O & L\\
	r & q & e & k & i & C & V & I & Q & Z\\
	u & v & h & {\bf p} & s & R & {\bf M} & A & G & U\\
	a & l & b & z & n & F & W & Y & H & S\\
	g & c & o & f & t & B & N & D & X & K\\
   \hline
	Q & T & B & L & E & v & q & i & p & g\\
	Z & H & N & D & X & s & t & u & o & h\\
	P & M & I & Y & C & n & r & d & x & y\\
	V & S & K & {\bf W} & O & b & {\bf l} & w & m & f\\
	U & A & F & R & G & c & z & k & a & e\\
   \hline
   \end{tabular}
   \caption{Four square cipher}
   \label{Four-Square-Cipher-table-reference}
   \end{center}
   \end{table}%

@@

   \begin{table}[h]
   \begin{center}
   \begin{tabular}{|c|c|c|c|c|}
   \hline
	Plaintext:  & {\tt the} & {\tt alphabet} & {\tt is} & {\tt changing}\\
   \hline
	Key:        & {\tt KEY} & {\tt KEYKEYKE} & {\tt YK} & {\tt EYKEYKEY}\\
   \hline
	Ciphertext: & {\tt DLC} & {\tt KPNREZOX} & {\tt GC} & {\tt GFKRESRE}\\
   \hline
   \end{tabular}  
   \end{center}

@@

   \begin{table}[h]
   \begin{center}
   \begin{tabular}{|c|c|c|c|c|}
   \hline
   Plaintext:  & {\tt the} & {\tt alphabet} & {\tt is} & {\tt changing}\\
   \hline
   Key:        & {\tt KEY} & {\tt THEALPHA} & {\tt BE} & {\tt TISCHANG}\\
   \hline
   Ciphertext: & {\tt DLC} & {\tt TSTHLQLT} & {\tt JW} & {\tt VPSPNIAM}\\
   \hline
   \end{tabular}  
   \caption{Autokey}	
   \label{Autokey-table-reference}
   \end{center} 
   \end{table}

@@

   alphabet: KEYWORDABCFGHILMNPSTUVXZ\\
   \begin{table}[h]
   \begin{center}
   \begin{tabular}{|c||r@{\:}r@{\:}r@{\:}|r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}|r@{\:}r@{\:}|r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}|}
   \hline
	Plaintext: & t & h & e & a & l & p & h & a & b & e & t & i & s & c & h & a & n & g & i & n & g\\
	Numbering: & 1 & 2 & 3 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 3 & 4 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11\\
	Ciphertext: & U & L & O & C & P & V & P & I & M & C & O & N & X & I & P & I & Z & T & X & Y & X\\
   \hline
   \end{tabular}  
   \caption{Ragbaby}
   \end{center} 
   \end{table}
   \end{itemize}

@@

	\begin{table}[h]
	\begin{center}
	\begin{tabular}{|c|ccccc|}
	\hline
		  & 2 & 4 & 5 & 1 & {\bf 3}\\
	\hline
		1 & K & E & Y & W & O\\
		{\bf 4} & R & D & A & B & {\bf C}\\
		2 & F & G & H & I & L\\
		3 & M & N & P & Q & S\\
		5 & T & U & V & X & Z\\
	\hline
	\end{tabular}  
	\end{center}

@@

   \begin{table}[h]
   \begin{center}
   \begin{tabular}{|ccccccccc|ccc|c|}
   \hline	
	1 & 2 & {\bf 3} & 4 & 5 & 6 & 7 & 8 & 9 &   &   &   &  \\
   \hline

@@

   \begin{table}[h]
   \begin{center}
   \begin{tabular}{|ccccccccccc|}
   \hline
	K & E & Y & & E & K & Y & & E & K & Y\\
   \hline
	c & o & m & & o & c & m & & S & M & K\\
	b & i & n & & i & b & n & & M & L & L\\
	i & n & g & & n & i & g & & R & S & E\\
	s & u & b & & u & s & b & & Y & C & Z\\
	s & t & i & & t & s & i & & X & C & G\\
	t & u & t & & u & z & t & & Y & J & R\\
	i & o & n & & o & i & n & & S & S & L\\
	a & n & d & & n & a & d & & R & K & B\\
	t & r & a & & r & t & a & & V & D & Y\\
	n & s & p & & s & n & p & & W & X & N\\
	o & s & i & & s & o & i & & W & Y & G\\
	t & i & o & & i & t & o & & M & D & N\\
   \hline
   \end{tabular}
   \caption{Nicodemus}
   \label{Nicodemus-table-reference}
   \end{center}
   \end{table}

@@

\item {\bf Card games} \cite{pp:Savard1999}: The key is created by means of a
   pack of cards and rules that are agreed upon in advance. All methods
   mentioned in this paragraph are designed as paper- and pencil methods,
   i.e. they are applicable without electronic aid. A pack of cards is
   unsuspicious to outsiders, shuffling the deck provides a certain amount
   of coincidence, cards can be transformed into numbers easily and a
   transposition cipher can be carried out without any further aid.
   \begin{itemize}
      \item {\bf Solitaire (Bruce Schneier)\footnotemark}
         \footnotetext{%
         %This cipher will be implemented in CrypTool in a future release.}
         In CrypTool\index{CrypTool} you can call this method under the menu
         {\bf Crypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
         Solitaire}.}
         \cite{pp:Schneier1999}\index{Solitaire}: 
         Sender and receiver have to own a deck of cards shuffled in the
         same manner. A key stream is generated that has to consist of as 
         many characters as the message to be encrypted.

@@

\item {\bf VIC cipher} \cite{pp:Savard1999}: This is a highly complicated but
   relatively secure paper- and pencil method. It has been developed and
   applied by Soviet spies. Amongst other things, the user had to create
   ten pseudo-random numbers out of a date, the first words of a sentence
   and any five-digit number. A straddling checkerboard is part of the
   encryption, too. A detailed description can be found under
   \cite{pp:Savard1999}.

@@

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\newpage
\section{Examples using Sage}
\label{PaP_Sage_samples}
\index{Sage!Code examples}
\index{Sage}

In the following section some classic ciphers are implemented using the open source computer
algebra system Sage. The code was tested with Sage version 4.1.
All ciphers are explained in the ``\nameref{Kapitel_PaperandPencil}'' chapter
(chap. no. \ref{Kapitel_PaperandPencil}).

@@

\noindent To make the sample code\footnote{%
     Further examples with Sage can be found e.g. at:\\
     \href{http://www.sagemath.org/doc/constructions/linear_codes.html\#classical-ciphers}{\tt
           http://www.sagemath.org/doc/constructions/linear\_codes.html\#classical-ciphers}
                                          }
easier to understand, we used the structure and the naming conventions
shown in the graphics below:
\begin{itemize}
  \item Encryption consists of the two steps encoding and enciphering.
  \item Encoding adapts the letters in the given plaintext P to the case defined in the
        given alphabet, and all non-alphabet characters are filtered out.
  \item Enciphering creates the ciphertext C.
  \item Decryption also consists of two steps: deciphering and decoding.
  \item Decoding is only necessary if the symbols in the alphabet are no ASCII characters.
\end{itemize}

\begin{figure}[h]
\begin{center}
\includegraphics[scale=0.5]{figures/encryption-decryption-en}
\caption{Structure and naming convention of the Sage cipher code examples} 
\label{XXX}
\end{center}
\end{figure}

@@

\bibitem[Schmeh2007]{pp:Schmeh2007}  \index{Schmeh 2007}
        Klaus Schmeh, \\
        {\em Codeknacker gegen Codemacher. Die faszinierende Geschichte der Verschl\"usselung.},\\ 
        W3L Verlag Bochum, 2. Auflage 2007.\\
	This is most current among the books dealing in a comprehensive manner with the history of cryptology. It contains a small collection of solved and unsolved crypto riddles.

>>>>

\chapter{Paper and Pencil Encryption Methods}
\label{Kapitel_PaperandPencil}
(Christine St\"otzel, April 2004; Updates: B.+C. Esslinger, June 2005; Updates Minh Van Nguyen and B. Esslinger, July 2009)
\index{Paper and pencil methods}

@@

The first paper and pencil methods already arose about 3000 years ago, but new
procedures were developed during the past century, too. All paper and pencil 
methods are a matter of symmetric methods\index{Encryption!symmetric}. 
Even the earliest encryption algorithms use the basic principles such as 
transposition, substitution, block construction and their combinations. 
Hence it is worthwhile to closely consider this ``ancient'' methods especially
under didactic aspects.

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}}
	  & n &   & x &   & m &   & l &   & o &   & t &   & a &   & s &   & o &   & i &   & i &   & n \\
	a &   & e &   & a &   & p &   & e &   & f &   & r &   & n &   & p &   & s &   & t &   & o &   \\
\end{tabular}
\caption{Rail Fence cipher}
\end{center} 
\end{table}

@@

   In the example for a turning grille in the following table you can write
   4 times 16 characters of the cleartext on a piece of paper:
\begin{table}[ht]
\begin{center}
\begin{tabular}{|cccc|cccc|}
\hline 	
	O & - & - & - & - & O & - & - \\
	- & - & - & O & O & - & - & O \\
	- & - & - & O & - & - & O & - \\
	- & - & O & - & - & - & - & - \\
\hline 	
	- & - & - & - & O & - & - & - \\
	O & - & O & - & - & - & O & - \\
	- & O & - & - & - & - & - & O \\
	- & - & - & O & O & - & - & - \\
\hline
\end{tabular}  
\caption{8x8 turning grille} 
\end{center}   
\end{table}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline 	
	K & E & Y \\
\hline
	a & n & e \\
	x & a & m \\
	p & l & e \\
	o & f & t \\
	r & a & n \\
	s & p & o \\
	s & i & t \\
	i & o & n \\
\hline
\end{tabular}
\caption{Simple columnar transposition}
\end{center} 
\end{table}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|c|c|c|c|c|}
\hline 	
	C & O & L & U & M & N\\
	1 & 5 & 2 & 6 & 3 & 4\\
\hline
	a &   &   &   &   &  \\
	n & e & x &   &   &  \\
	a & m & p & l & e &  \\
	o & f & t & r & a & n\\
	s & p &   &   &   &  \\
	o & s & i & t &   &  \\
	i & o & n &   &   &  \\
\hline
\end{tabular}
\caption{Columnar transposition (General Luigi Sacco)}
\end{center} 
\end{table}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|ccccc||cc|ccccc|}
\hline 	
	  & W & O & R & D & S &   &   & D & O & R & S & W\\
\hline
	W & a & n & e & x & a &   & D & s & p & o & i & s\\
	O & m & p & l & e & o &   & O & e & p & l & o & m\\
	R & f & t & r & a & n &   & R & a & t & r & n & f\\
	D & s & p & o & s & i &   & S & n & i & o & - & t\\
	S & t & i & o & n & - &   & W & x & n & e & a & a\\
\hline
\end{tabular}  
\caption[Nihilist transposition]{Nihilist transposition\footnotemark}
\end{center} 
\end{table}

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{|c|ccc|ccc|ccc|}
\hline 	
	  & K & {\bf E} & Y & {\bf E} & K & Y & {\bf E} & K & Y\\
\hline
	A & c & a & d & a & c & d & {\bf s} & a & a\\
	D & e & n & u & n & e & u & s & r & p\\
	X & s & i & s & i & s & s & i & f & i\\
	K & a & f & o & f & {\bf a} & o & u & l & o\\
	C & r & m & o & m & r & o & n & n & s\\
	W & f & c & o & c & f & o & k & t & g\\
	N & l & u & m & u & l & m & w & n & e\\
	S & n & a & r & a & n & r & d & o & o\\
	Y & t & r & a & r & t & {\bf a} & a & t & d\\
	{\bf E} & n & {\bf s} & p & {\bf s} & n & p & n & n & u\\
	D & o & s & i & s & o & i & i & i & s\\
	T & t & i & o & i & t & o & f & a & o\\	
	U & n & u & s & u & n & s & m & y & o\\
	B & i & n & g & n & i & g & c & r & o\\
	R & a & k & e & k & a & e & u & c & m\\
	G & y & w & o & w & y & o & a & e & r\\
	H & r & d & - & d & r & - & r & s & -\\
\hline
\end{tabular}  
\caption[Cadenus]{Cadenus\footnotemark}
\label{Cadenus-table-reference}
\end{center} 
\end{table}%

@@

   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|c|cccccccccc|}
   \hline 	
	  & 0 & 1 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9\\
   \hline
	  & K & - & - & E & Y & W & O & R & D & A\\
	1 & B & C & F & G & H & I & J & L & M & N\\
	2 & P & Q & S & T & U & V & X & Z & . & /\\
   \hline
   \end{tabular}  
   \caption{Straddling checkerboard with password ``Keyword''}
   \label{Straddling-Checkerboard-table-reference}
   \end{center} 
   \end{table}

@@

   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|c|ccccc|}
   \hline
        message              &  F   &   I   &   G   &   H   &   T     \\
   \hline
	ciphertext           & {\tt 00101} & {\tt 01000} & {\tt 00110} & {\tt 00111} & {\tt 10011} \\
	unsuspicious message & {\tt itisw} & {\tt arman} & {\tt thesu} & {\tt nissh} & {\tt ining} \\
   \hline 
        Baconian Cipher      & {\tt itIsW} & {\tt aRman} & {\tt thESu} & {\tt niSSH} & {\tt IniNG} \\
   \hline
   \end{tabular}  
   \caption{Baconian cipher}
   \label{Baconian-table-reference}
   \end{center} 
   \end{table}

@@

   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|c|c|c|c|c|}
   \hline
	K & E & Y & W & O\\
   \hline
	R & D & A & B & C\\
   \hline
	F & G & H & I & L\\
   \hline
	M & N & P & Q & S\\
   \hline
	T & U & V & X & Z\\
   \hline
   \end{tabular}
   \caption{5x5 Playfair matrix}
   \label{Playfair-table-reference}
   \end{center}
   \end{table}

@@

   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|ccccc|ccccc|}
   \hline
	d & w & x & y & m & E & P & T & O & L\\
	r & q & e & k & i & C & V & I & Q & Z\\
	u & v & h & {\bf p} & s & R & {\bf M} & A & G & U\\
	a & l & b & z & n & F & W & Y & H & S\\
	g & c & o & f & t & B & N & D & X & K\\
   \hline
	Q & T & B & L & E & v & q & i & p & g\\
	Z & H & N & D & X & s & t & u & o & h\\
	P & M & I & Y & C & n & r & d & x & y\\
	V & S & K & {\bf W} & O & b & {\bf l} & w & m & f\\
	U & A & F & R & G & c & z & k & a & e\\
   \hline
   \end{tabular}
   \caption{Four square cipher}
   \label{Four-Square-Cipher-table-reference}
   \end{center}
   \end{table}%

@@

   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|c|c|c|c|c|}
   \hline
	Plaintext:  & {\tt the} & {\tt alphabet} & {\tt is} & {\tt changing}\\
   \hline
	Key:        & {\tt KEY} & {\tt KEYKEYKE} & {\tt YK} & {\tt EYKEYKEY}\\
   \hline
	Ciphertext: & {\tt DLC} & {\tt KPNREZOX} & {\tt GC} & {\tt GFKRESRE}\\
   \hline
   \end{tabular}  
   \end{center}

@@

   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|c|c|c|c|c|}
   \hline
   Plaintext:  & {\tt the} & {\tt alphabet} & {\tt is} & {\tt changing}\\
   \hline
   Key:        & {\tt KEY} & {\tt THEALPHA} & {\tt BE} & {\tt TISCHANG}\\
   \hline
   Ciphertext: & {\tt DLC} & {\tt TSTHLQLT} & {\tt JW} & {\tt VPSPNIAM}\\
   \hline
   \end{tabular}  
   \caption{Autokey}	
   \label{Autokey-table-reference}
   \end{center} 
   \end{table}

@@

   alphabet: KEYWORDABCFGHILMNPSTUVXZ\\
   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|c||r@{\:}r@{\:}r@{\:}|r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}|r@{\:}r@{\:}|r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}r@{\:}|}
   \hline
	Plaintext: & t & h & e & a & l & p & h & a & b & e & t & i & s & c & h & a & n & g & i & n & g\\
	Numbering: & 1 & 2 & 3 & 2 & 3 & 4 & 5 & 6 & 7 & 8 & 9 & 3 & 4 & 4 & 5 & 6 & 7 & 8 & 9 & 10 & 11\\
	Ciphertext: & U & L & O & C & P & V & P & I & M & C & O & N & X & I & P & I & Z & T & X & Y & X\\
   \hline
   \end{tabular}  
   \caption{Ragbaby}
   \end{center} 
   \end{table}
   \end{itemize}

@@

	\begin{table}[ht]
	\begin{center}
	\begin{tabular}{|c|ccccc|}
	\hline
		  & 2 & 4 & 5 & 1 & {\bf 3}\\
	\hline
		1 & K & E & Y & W & O\\
		{\bf 4} & R & D & A & B & {\bf C}\\
		2 & F & G & H & I & L\\
		3 & M & N & P & Q & S\\
		5 & T & U & V & X & Z\\
	\hline
	\end{tabular}  
	\end{center}

@@

   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|ccccccccc|ccc|c|}
   \hline	
	1 & 2 & {\bf 3} & 4 & 5 & 6 & 7 & 8 & 9 &   &   &   &  \\
   \hline

@@

   \begin{table}[ht]
   \begin{center}
   \begin{tabular}{|ccccccccccc|}
   \hline
	K & E & Y & & E & K & Y & & E & K & Y\\
   \hline
	c & o & m & & o & c & m & & S & M & K\\
	b & i & n & & i & b & n & & M & L & L\\
	i & n & g & & n & i & g & & R & S & E\\
	s & u & b & & u & s & b & & Y & C & Z\\
	s & t & i & & t & s & i & & X & C & G\\
	t & u & t & & u & z & t & & Y & J & R\\
	i & o & n & & o & i & n & & S & S & L\\
	a & n & d & & n & a & d & & R & K & B\\
	t & r & a & & r & t & a & & V & D & Y\\
	n & s & p & & s & n & p & & W & X & N\\
	o & s & i & & s & o & i & & W & Y & G\\
	t & i & o & & i & t & o & & M & D & N\\
   \hline
   \end{tabular}
   \caption{Nicodemus}
   \label{Nicodemus-table-reference}
   \end{center}
   \end{table}

@@

\item {\bf Card games} \cite{pp:Savard1999}: The key is created by means of a
   pack of cards and rules that are agreed upon in advance. All methods
   mentioned in this paragraph are designed as paper and pencil methods,
   i.e. they are applicable without electronic aid. A pack of cards is
   unsuspicious to outsiders, shuffling the deck provides a certain amount
   of coincidence, cards can be transformed into numbers easily and a
   transposition cipher can be carried out without any further aid.
   \begin{itemize}
      \item {\bf Solitaire (Bruce Schneier)\footnotemark}
         \footnotetext{%
         %This cipher will be implemented in CrypTool in a future release.}
         In CrypTool\index{CrypTool} you can call this method under the menu
         {\bf Crypt/Decrypt \textbackslash{} Symmetric (classic) \textbackslash{} 
         Solitaire}.}
         \cite{pp:Schneier1999}\index{Solitaire}: 
         Sender and receiver have to own a deck of cards shuffled in the
         same manner. A key stream is generated that has to consist of as 
         many characters as the message to be encrypted.

@@

\item {\bf VIC cipher} \cite{pp:Savard1999}: This is a highly complicated but
   relatively secure paper and pencil method. It has been developed and
   applied by Soviet spies. Amongst other things, the user had to create
   ten pseudo-random numbers out of a date, the first words of a sentence
   and any five-digit number. A straddling checkerboard is part of the
   encryption, too. A detailed description can be found under
   \cite{pp:Savard1999}.

@@

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\newpage
\section{Appendix: Examples using Sage}
\label{PaP_Sage_samples}
\index{Sage!Code examples}
\index{Sage}

In the following section some classic ciphers are implemented using the open source computer
algebra system Sage. The code was tested with Sage version 4.1.
All ciphers are explained in chapter~\ref{Kapitel_PaperandPencil}
(``\nameref{Kapitel_PaperandPencil}'').

@@

\noindent To make the sample code\footnote{%
     Further examples with Sage can be found e.g. at:\\
     \href{http://www.sagemath.org/doc/constructions/linear_codes.html\#classical-ciphers}{\tt
           http://www.sagemath.org/doc/constructions/linear\_codes.html\#classical-ciphers}
                                          }
easier to understand, we used the structure and the naming conventions
shown in the graphics below:
\begin{itemize}
  \item Encryption consists of the two steps encoding and enciphering.
  \begin{itemize}
    \item Encoding adapts the letters in the given plaintext P to the case defined in the
          given alphabet, and all non-alphabet characters are filtered out.
    \item Enciphering creates the ciphertext C.
  \end{itemize}
  \item Decryption also consists of two steps: deciphering and decoding.
  \begin{itemize}
    \item Decoding is only necessary if the symbols in the alphabet are no ASCII characters.
  \end{itemize}
\end{itemize}

\begin{figure}[ht]
\begin{center}
\includegraphics[scale=0.5]{figures/encryption-decryption-en}
\caption{Structure and naming convention of the Sage cipher code examples} 
\label{XXX}
\end{center}
\end{figure}

@@

\bibitem[Schmeh2007]{pp:Schmeh2007}  \index{Schmeh 2007}
        Klaus Schmeh, \\
        {\em Codeknacker gegen Codemacher. Die faszinierende Geschichte der Verschl\"usselung.},\\ 
        W3L Verlag Bochum, 2. Auflage 2007.\\
	This is most current among the books dealing in a comprehensive manner
	with the history of cryptology. It contains a small collection of solved
	and unsolved crypto riddles. One of the challenges deals with a double
	column transposition using two long keys, which are differently in addition.
